\ripple: Differentiable Waveforms for Gravitational Wave Data Analysis

### 2023-01-18 src/tex/ms.tex
Edits

\AZ{Add refs here:}
There are a variety of gravitational waveforms currently used in analysis pipelines.
They are generally structured into different families, the most common of which are: the effective-one-body (EOB), the phenomenological inspiral-merger-ringerdown (IMRPhenom), and numerical relativity surrogate (NRsurrogate).
Of these, the IMRPhenom family is most well suited for an AD implementation in \jax.
In particular, their closed form expressions and fixed size frequency grids make a \jax implementation that complies with the constraints of JIT compilation simple.
\AZ{to fix:} EOB waveforms on the other hand require dynamic frequency grids and are therefore tricky to implement in \jax.
\AZ{What does this mean? I don't understand why EOB needs dynamic frequency grids. We should clarify, and we can also mention if \jax is well-suited to frequency-domain ROM implementations of EOB, and reference those.} 
In addition, EOB waveforms use hypergeometric functions which have limited support in \jax.
\AZ{rephrase this in a more positive fashion:}
NRsurrogate waveforms have no fundamental issue but require the user to download relatively large data files, making their use cumbersome in practice.\te{Max/Kaze, please add to/change this if you think its incorrect}

@@

Note that in General Relativity (GR) the total mass $M$ of \AZ{a binary black hole simply serves as an overall scale for the system, so that the frequency evolution can be trivially rescaled.
The total mass only impacts the match because the chosen PSD and frequency limits fix a reference scale.}
\sout{the system is simply an overall scaling of the waveforms frequency evolution and is therefore not a truly independent parameter.}
If we instead use a flat PSD in Eq.~\ref{eq:inner_prod} and \AZ{rescale the frequency grid to units of $Mf$}, \sout{correctly,} all dependence with $M$ seen in Fig.~\ref{fig:match} vanishes. 
\AZ{Revisit this after defining PSD above:} 
This figure instead illustrates a more realistic PSD where at low masses the waveform signal-to-noise is dominated by the inspiral but at high masses it is dominated by the merger.

@@

The primary arguement of this paper is that waveform derivatives will also be highly vauable to data analysis tasks. 
AD provides two big advantages when it comes to evaluating derivatives compared to numerical differentiation.
First, the accuracy of derivatives from AD are signficantly more stable than finite difference methods.
In particular, finite differences suffer from both rounding and truncation errors, meaning that the user is required to \textit{tune} the width over which the difference is taken.
On the other hand, AD produces machine precision derivatives with no tuninig.
Second, AD scales favourably with the dimensionality of the function.
In particular, for every input dimension, $D$, added one would need to evaluate the function at least $2D$ times to calcuate finite difference derivatives for all input parameters.
For reverse mode AD, one only needs two function calls to evaluate the derivative of all input parameters, regardless of dimenison.\footnote{
    Note that although the number of function calls is small, reverse mode AD does add memory overhead.
    We've not found this to be limiting in any of the situations tested so far.
}
Since the parameter space of GWs in GR has $\mathcal{O}(10)$ dimensons, the speed of derivative evaluation is less crucial than the stability.
However, this might change for waveforms in beyond GR models where many more parameters are added.

@@

Previously in the construction of IMRPhenomD \citep{Khan:2015jqa}, waveform
coefficients are fitted in segments with amplitude and phase separated.
Furthermore, IMRPhenomD is divided in three fitting segments, inpsiral, merger and ringdown.
Each of these segments has their own set of fitting coefficients.
After obtaining the fitting coefficient for individual segments, they are then
"stitched" together such that they are continuous in the first derivative.
The process of stitching introduces some addition
inaccuracy in the waveform model, as the connections would affect the originally
fitted segments.
The coefficients of the current IMRPhenomD implementation are tuned in sub-sets
of parameters instead of all together. This means the tuning process ignore the
correlation between different sub-sets of parameters, so the best-fitted
solution obtained by fitting sub-sets of parameters may not be the global
optimum compared to fitting all coefficients at once. By jointly
fitting all coefficients, the correlation in fitting individual and connecting
different segments can be accounted during the optimization. Therefore, there is
potential improvement of the accuracy by fitting all coefficients at once.

@@

Moreover, each error calculation (including computing the Fisher matrices for each detector and the inversion procress) is substantially faster.
In particular, we find that after compilation each error calculation takes approximately half a second on a single computing core.
\texttt{GWbench}~\citep{Borhanian:2020ypi}, on the other hand, takes $\mathcal{O}$(minutes) for each Fisher calculation using the same detector setup and frequency grid.
This factor of over 100 speed up is substantial considering the fact that a single core evaluation of the \ripple waveform is slower than the \lalsuite call.
As discussed above, performance can be further improved by utilising hardware acceleration such as parallel GPU processing.
AD therefore represents a fast and accurate way of performing population level analyses, and should be utilized for testing the capabilities of next generation detectors.

>>>>

\AZ{Add refs here:}
There are a variety of gravitational waveforms currently used in analysis pipelines.
They are generally structured into different families, the most common of which are: the effective-one-body (EOB), the phenomenological inspiral-merger-ringerdown (IMRPhenom), and numerical relativity surrogate (NRsurrogate).
%Of these, the IMRPhenom family is most well suited for an AD implementation in \jax.
\AZ{Of these, the IMRPhenom family serves as a natural starting point for an AD implementation in \jax.
Models like the non-precessing IMRPhenomD model studied here and the precessing, higher-mode model IMRPhenomXPHM~\citep{} are written in frequency domain using closed-form expressions.
This makes a}
\sout{In particular, their closed form expressions and fixed size frequency grids make a }
\jax implementation that complies with the constraints of JIT compilation simple.
\AZ{to fix:} EOB waveforms on the other hand require dynamic frequency grids and are therefore tricky to implement in \jax.
\AZ{What does this mean? I don't understand why EOB needs dynamic frequency grids. We should clarify, and we can also mention if \jax is well-suited to frequency-domain ROM implementations of EOB, and reference those.} 
In addition, EOB waveforms use hypergeometric functions which have limited support in \jax.
\AZ{rephrase this in a more positive fashion:}
NRsurrogate waveforms have no fundamental issue but require the user to download relatively large data files, making their use cumbersome in practice.\te{Max/Kaze, please add to/change this if you think its incorrect}

@@

Note that in General Relativity (GR) the total mass $M$ of \AZ{a binary black hole simply serves as an overall scale for the system, so that the frequency evolution can be trivially rescaled.
The total mass only impacts the match because the chosen PSD and frequency limits fix a reference scale.}
\sout{the system is simply an overall scaling of the waveforms} 
\sout{frequency evolution and is therefore not a} \sout{truly independent parameter.}
If we instead use a flat PSD in Eq.~\ref{eq:inner_prod} and \AZ{rescale the frequency grid to units of $Mf$}, \sout{correctly,} all dependence with $M$ seen in Fig.~\ref{fig:match} vanishes. 
\AZ{Revisit this after defining PSD above:} 
This figure instead illustrates a more realistic PSD where at low masses the waveform signal-to-noise is dominated by the inspiral but at high masses it is dominated by the merger.

@@

The primary argument of this paper is that waveform derivatives will also be highly valuable to data analysis tasks. 
AD provides two big advantages when it comes to evaluating derivatives compared to numerical differentiation.
First, the accuracy of derivatives from AD are significantly more stable than finite difference methods.
In particular, finite differences suffer from both rounding and truncation errors, meaning that the user is required to \textit{tune} the width over which the difference is taken.
On the other hand, AD produces machine precision derivatives with no tuning.
Second, AD scales favorably with the dimensionality of the function.
In particular, for every input dimension, $D$, added one would need to evaluate the function at least $2D$ times to calculate finite difference derivatives for all input parameters.
For reverse mode AD, one only needs two function calls to evaluate the derivative of all input parameters, regardless of dimenison.\footnote{
    Note that although the number of function calls is small, reverse mode AD does add memory overhead.
    We've not found this to be limiting in any of the situations tested so far.
}
Since the parameter space of GWs in GR has $\mathcal{O}(10)$ dimensions, the speed of derivative evaluation is less crucial than the stability.
However, this might change for waveforms in beyond GR models where many more parameters are added.

@@

Previously in the construction of IMRPhenomD \citep{Khan:2015jqa}, waveform
coefficients are fitted in segments with amplitude and phase separated.
Furthermore, IMRPhenomD is divided in three fitting segments, inspiral, merger and ringdown.
Each of these segments has their own set of fitting coefficients.
After obtaining the fitting coefficient for individual segments, they are then
"stitched" together such that they are continuous in the first derivative.
The process of stitching introduces some addition
inaccuracy in the waveform model, as the connections would affect the originally
fitted segments.
The coefficients of the current IMRPhenomD implementation are tuned in sub-sets
of parameters instead of all together. This means the tuning process ignore the
correlation between different sub-sets of parameters, so the best-fitted
solution obtained by fitting sub-sets of parameters may not be the global
optimum compared to fitting all coefficients at once. By jointly
fitting all coefficients, the correlation in fitting individual and connecting
different segments can be accounted during the optimization. Therefore, there is
potential improvement of the accuracy by fitting all coefficients at once.

@@

Moreover, each error calculation (including computing the Fisher matrices for each detector and the inversion process) is substantially faster.
In particular, we find that after compilation each error calculation takes approximately half a second on a single computing core.
\texttt{GWbench}~\citep{Borhanian:2020ypi}, on the other hand, takes $\mathcal{O}$(minutes) for each Fisher calculation using the same detector setup and frequency grid.
This factor of over 100 speed up is substantial considering the fact that a single core evaluation of the \ripple waveform is slower than the \lalsuite call.
As discussed above, performance can be further improved by utilising hardware acceleration such as parallel GPU processing.
AD therefore represents a fast and accurate way of performing population level analyses, and should be utilized for testing the capabilities of next generation detectors.
